---
title: "Cannabis Sales in Colorado Counties"
author: "Lisa Marie Pritchett"
date: "January 23, 2018"
output:
  html_document:
    keep_md: yes
  pdf_document: default
header-includes: \usepackage{bbm}
---

```{r setup, include=FALSE, warning=F, message=F,}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
#opts_knit$set(root.dir = "~/data/")
#Clear environment
rm(list=ls())
library(knitr)
library(dplyr)
library(stringr) 
library(tidyr)
library(ggplot2)
library(XLConnect) # reads excel
library(leaflet) # creates maps
library(prettydoc) # html formatting
library(caret)
source("http://peterhaschke.com/Code/multiplot.R")
```


The state of Colorado has had legal recreational cannabis usage since 2014. In that time, more thant 4 Million dollars of Cannabis has been sold across the state fueling the local economies. 

The state regularly publishes the monthly cannabis sales in each county, with only a few exceptions.
The exceptions are those counties where the sales data is from less than 3 tax-payers (store owners). In these cases, for privacy, the state combines these counties and reports the Sum of all No Report ("NR") Counties. 

I estimate the sales in the NR Counties with linear regression to predict the amount of sales based on the county population and number of stores. Future work can add more predictors such as presence of regulations and more demographic information. I use those predictions and the known Sum of NR Counties to arrive at an estimate for the monthly sales in each of the NR Counties.


### Get Supporting Data

First I get all the supporting data for the project. The _R_ scripts loaded below are available on my github at https://github.com/lisamariepritchett/CannabisSalesColorado. All the coded needed to download data from state websites, read it into _R_, and clean and processes it is provided.


```{r getdata,  cache=T}
source('data/get_county_populations.R')
source('data/get_county_centers.R')
source('data/get_stores.R')
```

The scripts return data regarding each county in Colorado and its population for each year between 2014 and 2017. I also download a list of all of the medical and recreational cannabis outlets across the state. Note this is a current list. Historical information on medical and retail outles is available at the state website, but it is published as pdf files which are particularly hard to scrape. 

Next I visualize the population density using a Leaflet plot:


```{r plotpopulation, out.width='100%'}
county_details <- inner_join(county_centers,county_populations, by="County")
county_details %>% leaflet() %>% addProviderTiles(providers$Stamen.Terrain) %>% 
        addCircles(radius=sqrt(county_details$Population*2000), weight=0.5,
                   color='blue',fillOpacity = .05,
                   label=paste(county_details$County,
                               round(county_details$Population/1000),
                               'K Population'))

DenverMetroArea <- c('Denver','Jefferson','Adams','Arapahoe','Douglas')
GreaterDenverArea <- c(DenverMetroArea, 'Boulder', 'El Paso')
FrontRange <- c(GreaterDenverArea, 'Larimer','Weld','Pueblo')
```

The area of each of the circles represnets the population of the county it is centered around. 
We can see Denver is the population center. 
Several counties representing the Denver Metro Area (DMA). 
The next largest counties Boulder and El Paso (Colorado Springs) are adjascent to the DMA, and make up the Greater Denver Area. 
The population is most dense along the Front Range spanning from Larimer County (Ft Collins) to Pueblo. The only county outside the front range with more than 100,000 people is Mesa (Grand Junction) at the western border.

Next we can plot the density of cannabis stores across the state. Below a cluster map is used to show the density of licensed medical and retail outlets across the state. 

```{r storecluster, out.width='100%'}
combined_stores %>% leaflet() %>% addProviderTiles(providers$CartoDB) %>% 
        addMarkers(clusterOptions = markerClusterOptions(), 
                   label = paste(combined_stores$zip_code,
                                 combined_stores$LICENSEE,
                                 'Medical:', 
                                 !is.na(combined_stores$med.license),
                                 'Retail:',
                                 !is.na(combined_stores$retail.license)))

```

This interactive visualization illustrates the density of cannabis retail outlets across the state. I have combined medical and recreational outlets counting each unique address only once (even if it was licensed for both medical and retail sale). Individual markers point to the center of the zipcode where the store is located. Labels indicate the zipcode, licensee, and whehter the outlet is licensed for medical and retial sales. 

There are `r filter(county_n.stores,kind=='Combined') %>% select(n.stores) %>% sum()` licensed cannabis outlets across the state (a few not shown due to geocoding problem with their zipcodes). Of those, 
`r sum(filter(county_n.stores,County=='Denver') %>% select(n.stores))` or 
 `r round(sum(filter(county_n.stores,County=='Denver') %>% select(freq)*100))`% are within the County of Denver,
 `r round(sum(filter(county_n.stores,County %in% DenverMetroArea) %>% select(freq)*100))`% are within the Denver Metro Area,
 `r round(sum(filter(county_n.stores,County %in% GreaterDenverArea) %>% select(freq)*100))`% are within the Greater Denver Area, and
  `r round(sum(filter(county_n.stores,County %in% FrontRange) %>% select(freq)*100))`% are along the Front Range.


### Get Cannabis Sales Data

The State of Colorado publishes its report of cannabis sales in an excel spreadsheet for each month so the data needs to be gathered across dozens of files. The state publishes the total amount of sales reported in each county except in counties where less than 3 taxpayers report sales, in that case they omit the report for the county (label it NR in published data) and combine it with the other NR couties to report the overall Sum of NR Counties. In the code at data/get_sales.R I replaced the "NR" values with the known Sum of NR Counties divided by the number of NR Counties. Thus filling in the missing data such that the totals will add up to the totals reported by the state.


```{r get_sales, cache=T}
source('data/get_sales.R')
```


### Join 

Now I combine the cannabis sales data and other supporting data. 

```{r join_data_sets}
## Join county details with cannabis sales 
cannabis.sales <- full_join(cannabis.sales,county_details, by = c("County","year"))
cannabis.sales <- full_join(cannabis.sales,county_n.stores, by=c('County','kind'))

## Join Store addresses with zipcode to provide geocode for stores
med_stores <- left_join(med_stores,co_zipcodes)
retail_stores <- left_join(retail_stores,co_zipcodes)
combined_stores <- left_join(combined_stores,co_zipcodes)
```


### Improve Estimates for NR Counties

I can use the data I have to make better predictions about the amount of sales in the NR Counties. So far I have replaced NR with the Sum of NR / Number of NR, thus each NR estimate is the average amongst the NR Counties for that month. This assumes that the stores in tiny San Juan County (poplation about 700) sells as much as the stores in larger El Paso County (population 700,000). Both Counties have only two recreational cannabis outlets and are regularly reported as NR.

I improve the estimates using a linear model where I predict sales as a function of number of stores, county population, and outlet kind (medical/recreational). The code below shows a plot of the 


#### Fit Model 

```{r trainlm, out.width='70%', fig.show='hold'}
# cases to fit a linear model to
data <- cannabis.sales[complete.cases(cannabis.sales),] %>% 
        filter(!is.Estimate, USD>0, n.stores>0)

# partition train and test sets
set.seed(419)
InTrain <- createDataPartition(data$USD,p = .7, list = F)
train <- data[InTrain,]
test <- data[-InTrain,]

train %>% 
        ggplot(aes(x=log(Population),color=log(n.stores), 
                   shape=is.medical, y=log(USD))) + 
        geom_point(alpha=.5,size=3) + 
        stat_smooth(method = 'lm', alpha=.5) 
        ylim(c(10,18))


sales.lmfit <- train(log(USD) ~  log(n.stores) + log(Population)  + is.medical, data=train, method='lm')
summary(sales.lmfit)
```


The plot  above shows the fit of the linear model. More predictor variables could be added to improve model accuracy. This simple model accounts for more than 86% of observed varaince in log sales. 

#### Predict Sales in Counties Not Reporting (NR)

```{r predictNR, out.width='50%', fig.show='hold'}
# Now Make Predictions for Sales in NR Counties.

# Filter only cases which were NR.I change n.stores to be no higher than 2 because I know all the counties had 2 or fewer tax payers when they were NR. Some are higer than 2 because I only have current store counts not historical store counts. 
NRCounties <- cannabis.sales %>% 
        filter(is.Estimate,  USD>0) %>% 
        mutate(n.stores=ifelse(n.stores>2,2,n.stores),
               n.stores=ifelse(n.stores<1,1,n.stores))

NRCounties$predict <- exp(predict(sales.lmfit,NRCounties))

NRCounties %>% ggplot(aes(x=log(Population), color=factor(n.stores), shape=is.medical, y=log(predict))) + 
        geom_point(size=3) +
        scale_color_manual(values = c('navy','cornflowerblue'))

NRCounties %>% ggplot(aes(x=Population, color=factor(n.stores),
                          shape=is.medical, y=predict)) + 
        geom_point(size=3)  +
        scale_color_manual(values = c('navy','cornflowerblue'))


```

The plots above show the predicted sales as a function of population, number of stores, and type. 
This model predicts that `r NRCounties$predict %>% sum(na.rm=T)/10^6` Million will be sold in these counties overall. 
However, the state reports that the actual total amount across these counties is  
`r NRCounties$USD %>% sum(na.rm=T) /10^6` Million USD. I correct this in the code below, keeping the porportion of sales between NR Counties as predicted by my model, but scaling so that the sum adds to the known totals.

```{r checksum}
 NRCounties <- NRCounties %>% group_by(date) %>% 
        mutate(USD.portion = predict / sum(predict),
               USD.newpredict = USD.portion * sum(USD)) %>% 
        arrange(desc(USD.portion))

NRCounties %>% group_by(date) %>% summarise(known.total=sum(USD),predicted.total=sum(USD.newpredict))

NRCounties <- NRCounties %>% mutate(USD=USD.newpredict) %>% 
        select(colnames(cannabis.sales))
```

The output above shows that the known totals and my transformed predicted totals for each of the first 10 months are exactly equal. In addition, the overall total now sums to the same figure reported by the state:
`r sum(NRCounties$USD)/10^6` Million USD.

Next I filter out my old estimates and replace them with these new estimates.

```{r replace}
# Replace in cannabis.sales 
# Filter out rows with old estimates
cannabis.sales <- cannabis.sales %>% filter(!is.Estimate)
cannabis.sales <- bind_rows(cannabis.sales,NRCounties)

# calculate perCapita sales and perStore sales
cannabis.sales <- cannabis.sales %>% mutate(
        perCapita=USD/Population,
        perStore = USD/n.stores) 

# reorder columns
cannabis.sales <- cannabis.sales %>% 
        select(County,County_ID,USD,perCapita,perStore,
               latitude,longitude,type,kind,year,
               date,is.medical,is.Estimate,
               Population,n.stores) %>% 
        as_tibble()
```

Now I'm ready to make plots showing the distribution of cannabis sales across the state.

### Make Plots










```{r USDmap, out.width='100%'}
cannabis.totals <- cannabis.sales %>% group_by(County,type,latitude,longitude) %>% 
        summarise(USD=sum(USD, na.rm = T),
                  perCapita=sum(perCapita, na.rm=T),
                  perStore=sum(perStore, na.rm=T))

myplotdata <- cannabis.totals %>% select(County,type,latitude,longitude,USD) %>% 
        spread(key=type,value=USD) %>% 
        mutate(Combined.NR = sum(medical.NR,recreational.NR,na.rm=T),
               CombinedNr_plus_med = sum(Combined.NR, medical.sales, na.rm=T),
               CombinedNr_plus_plus_rec = sum(CombinedNr_plus_med, recreational.sales,na.rm=T),
               label = paste(County, ':$', 
                             round((CombinedNr_plus_plus_rec-Combined.NR)/10^6),
                             ' Million Reported+$',
                             round(Combined.NR/10^6),
                             ' Million Estimated',
                             sep='')
        ) %>% 
        arrange(desc(CombinedNr_plus_plus_rec))
nocannabis <- county_centers[!(county_centers$County %in% cannabis.totals$County),]

radiusscale=1
myplotdata %>% leaflet() %>% addProviderTiles(providers$CartoDB) %>% 
        # Show the biggest circles 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_plus_rec*radiusscale),
                   color='#008b45',weight=1,fillOpacity = 0.25,
                   label=myplotdata$label
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_med*radiusscale),
                   color='#008b45',weight=0,fillOpacity = 0.25,
                   label=myplotdata$label                   
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$Combined.NR*radiusscale),
                   color='#8b0046',weight=0,fillOpacity = 0.25,
                   label = myplotdata$label
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$medical.NR*radiusscale),
                   color='black',weight=0,fillOpacity = 0.25,
                   label=myplotdata$label
        ) %>% 
        addCircles(lat=nocannabis$latitude,
                   lng=nocannabis$longitude,
                   radius=10,
                   color='red',  weight=0,
                   label=paste(nocannabis$County,'-'))
```
```{r USDbar, fig.width=7,fig.height=6}
cannabis.sales <- cannabis.sales %>%
        mutate(type=factor(type, levels = c('recreational.sales', 
                                            'medical.sales',
                                            'recreational.NR', 
                                            'medical.NR')))

denver <- cannabis.sales %>% filter(County == 'Denver') %>% 
        ggplot(aes(x='       Denver', y=USD/10^9, fill=type)) + 
        geom_bar(stat="identity") + 
        labs(title='Total Cannabis Sales in Colorado, Medical and Recreational By County',
             y='Billions USD',x='') +
        scale_fill_manual(values = alpha(c('seagreen3','springgreen4'), 0.7)) +
        scale_y_continuous(position = "right")+ coord_flip() +
        theme(legend.position = 'none') +
        ylim(0,2)

allothers <- cannabis.sales %>% arrange(County,type) %>% 
        filter(County != 'Denver') %>% 
        ggplot(aes(x=County, y=USD/10^6, fill=type)) + 
        geom_bar(stat="identity", position = 'stack') + 
        coord_flip() + ylab('Millions of USD') +
        ylim(c(0,450)) +
        scale_fill_manual(values = alpha(c('seagreen3','springgreen4','gray60','gray20'), 0.7))


multiplot(plotlist = list(denver,allothers), layout = matrix(c(1,2,2,2,2), nrow=5, byrow = T))

```


#### Per Capita

```{r perCapitamap,  out.width='100%'}
myplotdata <- cannabis.totals %>% select(County,type,latitude,longitude,perCapita) %>% 
        spread(key=type,value=perCapita) %>% 
        mutate(Combined.NR = sum(medical.NR,recreational.NR,na.rm=T),
               CombinedNr_plus_med = sum(Combined.NR, medical.sales, na.rm=T),
               CombinedNr_plus_plus_rec = sum(CombinedNr_plus_med, recreational.sales,na.rm=T),
               label = paste(County, ':$', 
                             round((CombinedNr_plus_plus_rec-Combined.NR)),
                             ' Reported+$',
                             round(Combined.NR),
                             ' Estimated',
                             sep='')
        ) %>% 
        arrange(desc(CombinedNr_plus_plus_rec))
nocannabis <- county_centers[!(county_centers$County %in% cannabis.totals$County),]

radiusscale=10^5
myplotdata %>% leaflet() %>% addProviderTiles(providers$CartoDB) %>% 
        # Show the biggest circles 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_plus_rec*radiusscale),
                   color='#0067C2',weight=1,fillOpacity = 0.25,
                   label=myplotdata$label
        ) %>% 
        
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_med*radiusscale),
                   color='#0067C2',weight=0,fillOpacity = 0.25,
                   label=myplotdata$label                   
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$Combined.NR*radiusscale),
                   color='#c25b00',weight=0,fillOpacity = 0.25,
                   label = myplotdata$label
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$medical.NR*radiusscale),
                   color='black',weight=0,fillOpacity = 0.25,
                   label=myplotdata$label
        ) %>% 
        addCircles(lat=nocannabis$latitude,
                   lng=nocannabis$longitude,
                   radius=10,
                   color='red',  weight=0,
                   label=paste(nocannabis$County,'-'))
```

```{r perCapita, fig.width=7,fig.height=6}
cannabis.sales %>% 
        ggplot(aes(x=County, y=perCapita, fill=type)) + 
        geom_bar(stat="identity", position = 'stack') + 
        coord_flip() +
        scale_fill_manual(values = alpha(c('steelblue1','steelblue4','gray60','gray20'), 0.7)) +
        labs(title='Total Sales per Capita in each County', y='USD')
```


#### Per Store
```{r perStoremap,  out.width='100%'}
myplotdata <- cannabis.totals %>% select(County,type,latitude,longitude,perStore) %>% 
        spread(key=type,value=perStore) %>% 
        mutate(Combined.NR = sum(medical.NR,recreational.NR,na.rm=T),
               CombinedNr_plus_med = sum(Combined.NR, medical.sales, na.rm=T),
               CombinedNr_plus_plus_rec = sum(CombinedNr_plus_med, recreational.sales,na.rm=T),
               label = paste(County, ':$', 
                             round((CombinedNr_plus_plus_rec-Combined.NR)/10^6),
                             ' Million Reported+$',
                             round(Combined.NR/10^6),
                             ' Million Estimated',
                             sep='')
        ) %>% 
        arrange(desc(CombinedNr_plus_plus_rec))
nocannabis <- county_centers[!(county_centers$County %in% cannabis.totals$County),]

radiusscale=20
myplotdata %>% leaflet() %>% addProviderTiles(providers$CartoDB) %>% 
        # Show the biggest circles 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_plus_rec*radiusscale),
                   color='#6F137E',weight=1,fillOpacity = 0.25,
                   label=myplotdata$label
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$CombinedNr_plus_med*radiusscale),
                   color='#6F137E',weight=0,fillOpacity = 0.25,
                   label=myplotdata$label                   
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$Combined.NR*radiusscale),
                   color='#227e13',weight=0,fillOpacity = .25,
                   label = myplotdata$label
        ) %>% 
        addCircles(lat=myplotdata$latitude, 
                   lng=myplotdata$longitude,
                   radius = sqrt(myplotdata$medical.NR*radiusscale),
                   color='black',weight=0,fillOpacity = .25,
                   label=myplotdata$label
        ) %>% 
        addCircles(lat=nocannabis$latitude,
                   lng=nocannabis$longitude,
                   radius=10,
                   color='red',  weight=0,
                   label=paste(nocannabis$County,'-'))

```



```{r perStore, fig.width=7,fig.height=6}
cannabis.sales %>% 
        ggplot(aes(x=County, y=perStore/10^6, fill=type)) + 
        geom_bar(stat="identity", position = 'stack') + 
        coord_flip() + 
        #ylim(c(0,450))# +
        scale_fill_manual(values = alpha(c('orchid','darkorchid4','gray60','gray20'), 0.7)) +
        labs(title='Total Sales perStore in each County', y='Millions of USD')

```


